---
title: "R Notebook"
output: html_notebook
---


```{r}
library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(PNWColors)
loc <-  pnw_palette("Mushroom", 6)
stage <-pnw_palette("Winter", 7)

meta.func <-  read.csv(here("00_data", "03_func_data", "func_metadata", "mastercog_phrog.csv"), row.names=1)

COG.master <- read.csv(here("00_data", "03_func_data", "func_metadata", "mastercog.csv"), header=TRUE)


#Spor <- read.csv(here("03_func", "01_fishertests", "01_AVGs", "unadj_pval", "AVG_SporBacil_Lifestyle_fisher_unadj.csv"), row.names=1)
Spor <- read.csv(here("03_func", "01_fishertests", "01_AVGs", "unadj_pval", "AVGConsvSpor_SporBacil_Lifestyle_fisher_unadj.csv"), row.names=1)

NoSpor <- read.csv(here("03_func", "01_fishertests", "01_AVGs", "unadj_pval", "AVGConsvSpor_NonSporBacil_Lifestyle_fisher_unadj.csv"), row.names=1)

all.fisher <- bind_rows(Spor, NoSpor)


all.fisher <- merge(all.fisher, COG.master, by="COGlet", all.x=TRUE, all.y=FALSE)




##this defeats the purpose of rbinding but is temporarily helpful for ddebugging code
#all.fisher <- subset(all.fisher, all.fisher$fisher_test=="AVG_SporBacil_Lifestyle")

#hmm<- subset(hmm, hmm$completeness >=
```


```{r}

#all.fisher <- bind_rows(inph.fisher, gsv.fisher, uhgv.fisher)

p.signif <- 1e-3

k_threshold <- 5

all.fisher <- subset(all.fisher, all.fisher$gene_name!="lytD" & all.fisher$gene_name!="nlpD")


all.fisher <-  all.fisher %>%  mutate(p.adj = signif (p.adjust(p_two_sided, method = "BH"),5), neg_log10_p = -log10(p.adj),) %>% 
  # mark significantly enriched genes
   mutate(significant = (p.adj < p.signif) & (prevalence > k_threshold)) 

sig.fisher <- subset(all.fisher, all.fisher$significant==TRUE)




```



```{r}
library(dplyr)
library(ggplot2)
# --- NEW: allow two independent fill scales ---
# install.packages("ggnewscale") if you don't have it
library(ggnewscale)

#sig.fisher <- subset(all.fisher, all.fisher$significant==TRUE & all.fisher$fisher_test=="SporBacil_Lifestyle")




#localization_order <-  c("Pre-division", "Both", "Mother cell", "Forespore", "Spore", "Unknown")
#localization_order <- rev(localization_order)
# (replace above with your real localization order)
#stage_order <- c("Initiation", "Early", "Middle", "Late", "Mother cell Lysis", "Germination",  "Unknown")
#stage_order <- rev(stage_order)



sig.fisher <- subset(sig.fisher, sig.fisher$COG.broad!="INFORMATION STORAGE AND PROCESSING" & sig.fisher$COG.broad!="PHAGE")

sig.fisher <- subset(sig.fisher, sig.fisher$Nog.abbrev!="Gen. function (R)" & sig.fisher$Nog.abbrev!="Unknown function (S)")

# build pathway_levels (make sure you use the correct factor for localization)
pathway_levels <- sig.fisher %>%
  distinct(Nog.new, Nog.abbrev, gene_name) %>%
  mutate(Nog.new, Nog.abbrev, gene_name) %>%
  arrange(Nog.new, Nog.abbrev, gene_name) %>%
  pull(gene_name)

pathway_levels <- sig.fisher %>%
  distinct(COG.broad, Nog.abbrev, gene_name) %>%
  mutate(COG.broad, Nog.abbrev, gene_name) %>%
  arrange(COG.broad, Nog.abbrev, gene_name) %>%
  pull(gene_name)

sig.fisher.bubble <- sig.fisher %>%
  mutate(gene_name = factor(gene_name, levels = pathway_levels))

broad <-  pnw_palette("Sailboat", 6)

#loc.rev <- pnw_palette("Shuksan2", 7)
cog.cat <- pnw_palette("Moth", 13)

#sig.fisher.bubble <- subset(sig.fisher.bubble, sig.fisher.bubble$sporulation_impact!="Unknown")

p <- ggplot(sig.fisher.bubble, aes(x = fisher_test, y = gene_name)) +
  geom_point(aes(size = plot_OR,
                # shape = direction,
                 color= direction))+
            # alpha = 0.7) +

  scale_color_manual(values = c(
    "Temperate" = "black",
    "Lytic" = "grey70",
    "Equal" = "grey70"
  )) +
#scale_shape_manual(values = c(
#  "Temperate" = 16,  #1 # open circle
#  "Lytic" = 16, #16     # filled circle
#  "Equal" = 1        # optional — open circle
#))+
  labs(
    x = "Dataset",
    y = "Gene",
    color = "Lifestyle Enriched In",
    size = "|log2(OR)|"
  ) +
  
  theme_minimal()  + facet_wrap(~dataset)
p

strip_df <- unique(select(sig.fisher.bubble, COG.broad, Nog.new, Nog.abbrev, gene_name))
strip_df$COG.broad <- as.factor(strip_df$COG.broad)
strip_df$Nog.new <- as.factor(strip_df$Nog.new)
strip_df$Nog.abbrev <- as.factor(strip_df$Nog.abbrev)
strip_df <- strip_df %>% mutate(xpos = as.numeric(COG.broad))

strip_x <- min(strip_df$xpos) - 0.8   # ← ONLY CHANGE (was max + offset)

strip_df <- strip_df %>%
  select(COG.broad, Nog.new, Nog.abbrev, gene_name) %>%
  distinct() %>%
  mutate(x = strip_x, xend = strip_x + 0.4)

# ------------------ MINIMAL NECESSARY CHANGES BELOW -----------------------
# create two different x positions (so the second tile bar doesn't overlap)
strip_df <- strip_df %>%
  mutate(
    x_local = x,                # left-most tile (keeps your original x)
    x_stage = x - 0.45          # second tile shifted further left
  )
# ... your code up to first geom_tile() is unchanged ...

p <- p +
  geom_tile(
    data = strip_df,
    aes(x = x_local, y = gene_name, fill = COG.broad),  # <- use x_local
    width = 0.4,
    height = 0.9,
    inherit.aes = FALSE
  ) +
  coord_cartesian(clip = "off") +
  scale_fill_manual(   # <-- added name here
    name = "Cog Category",
    values = broad
    #values = loc.rev
  )+  guides(fill = guide_legend(reverse = TRUE) )

p
ggsave(here("03_func", "figs", "ConsvSpor_AVGEnrichment_LifestylexBacillSpor_Bubble.png"), width = 11, height = 8)
```



# start a new fill scale so the second geom_tile can use fill = stage independently
p <- p + ggnewscale::new_scale_fill()

p <- p +
  geom_tile(
    data = strip_df,
    aes(x = x_stage, y = gene_name, fill = COG.broad),       # <- use x_stage
    width = 0.4,
    height = 0.9,
    inherit.aes = FALSE
  ) +
  coord_cartesian(clip = "off") +
  scale_fill_manual(   # <-- added name here
    name = "Broad COG Function",
    values = broad
  )+  guides(fill = guide_legend(reverse = TRUE) )

p
# Add legend and show — only set size guide so fill legends keep their scale names
all.gg <- p + guides(
  size = guide_legend(title = "Odds Ratio "),
  color = guide_legend(
    title = "Lifestyle Enriched In",
    override.aes = list(size = 5)   # increase legend point size
  )
)
all.gg
all.gg 

###ggsave(here("03_func", "figs", "SporEnrichment_SporBacillLifestyle_Bubble.png"), width = 11, height = 8)+ facet_wrap(~fisher_test)

##ggsave(here("03_func", "figs", "SporEnrichment_LifestylexBacillSpor_Bubble.png"), width = 11, height = 8)+ facet_wrap(~fisher_test)


```{r}
TEST <- ggplot(sig.fisher, aes(x = dataset, y = reorder(gene_name, plot_OR))) +
  geom_point(aes(size = plot_OR,
                 color = direction),
             alpha = 0.7) +

  scale_color_manual(values = c(
    "Temperate" = "#1f78b4",
    "Lytic" = "#e31a1c",
    "Equal" = "grey70"
  )) +
  
  labs(
    y = "Gene",
    color = "Enriched in",
    size = "|log2(OR)|"
  ) +
  
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
  ) #+ facet_wrap(~stage, scales = "free_y")

TEST
```

```{r}
library(ggplot2)
library(ggrepel)

loc <-  pnw_palette("Starfish", 6)
stage <-rev(pnw_palette("Sailboat", 8))

###visualize by dataset and lifestyle color

ggplot(sig.fisher, aes(x =plot_OR, y = neg_log10_p)) +
  geom_point(aes(size = plot_OR,
                 color = direction),
             alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  scale_size_continuous(name = "|log2(OR)|") +
  scale_color_manual(values = c(
    "Temperate" = "#1f78b4",
    "Lytic" = "#e31a1c",
    "Equal" = "grey70"
  )) +
  labs(
    x = "log2(Odds Ratio)",
    y = "-log10(BH-adjusted p-value)",
    color = "Enriched in"
  ) +
  theme_minimal() +  # Add gene labels
  geom_text_repel(aes(label = gene_name),
                  size = 6,
                  max.overlaps = Inf)  + facet_grid(dataset~direction) 



library(ggplot2)
library(ggrepel)

###visualize by dataset and lifestyle color

ggplot(sig.fisher, aes(x =plot_OR, y = neg_log10_p)) +
  geom_point(aes(#size = plot_OR,
                 color = Nog.new),
             alpha = 0.7, size=6) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  scale_size_continuous(name = "|log2(OR)|") +
  labs(
    x = "log2(Odds Ratio)",
    y = "-log10(BH-adjusted p-value)",
    color = "Enriched in"
  ) +  # scale_color_manual(   # <-- added name here
   # name = "Location in Spore",
  #  values = rev(stage) ) +
  theme_minimal() +  # Add gene labels
  geom_text_repel(aes(label = gene_name),
                  size = 4,
                  max.overlaps = Inf) +facet_wrap(~direction)  + facet_grid(dataset~direction) 

##ggsave(here("03_func", "figs", "SporEnrichment_SporBacillLifestyle_PvalBubble.png"), width = 11, height = 8)


```





```{r}
ggplot(sig.fisher, aes(x = dataset, y = reorder(gene_name, plot_OR))) +
  geom_point(aes(size = plot_OR,
                 color = direction),
             alpha = 0.7) +

  scale_color_manual(values = c(
    "Temperate" = "#1f78b4",
    "Lytic" = "#e31a1c",
    "Equal" = "grey70"
  )) +
  
  labs(
    y = "Gene",
    color = "Enriched in",
    size = "|log2(OR)|"
  ) +
  
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
  ) #+ facet_wrap(~stage, scales = "free_y")

```

