
```{r}
library(here)
library(tidyverse)
library(PNWColors)
library(knitr)
library(cowplot)
library(ggrepel)


here::here()
data_dir <- here("03_func", "inphared")
meta_dir <- here("03_func", "03_metadata")
raw_dir <- here("03_func", "inphared", "03_inph_raw")
figs_dir <- here("03_func", "inphared", "figs")


pal2.sail <- pnw_palette(name="Sailboat",n=2,type="discrete")

pal2.shuk <- pnw_palette(name="Shuksan2",n=3,type="discrete")

pal2 <- c("#015b58", "#5962b5")
pal2.flip <- c("#5962b5", "#015b58")

pal2.star <- c("#24492e", "#89689d")


dan.enrich <- read.csv(here(meta_dir, "viral_sporGenes.csv"))
dan.enrich <- dan.enrich[,c(1:2)]


kegg.id <- read.csv(here(meta_dir, "UniprotSporulationList_Master.csv"))
#kegg.id <- kegg.id[,c(3:5)]
kegg.id[is.na(kegg.id)] <- "MISSING"
kegg.id <- subset(kegg.id, kegg.id$gene_id.ko!="MISSING")


hmm <- read.csv(here(data_dir, "inphared_hmm_comp2.csv"), row.names=1 ) ##d_amg
hmm$comp3 <- hmm$Nog.abbrev
#hmm$comp3 <- ifelse(hmm$sporulation_hit!="no_hit", "Sporulation", hmm$comp3)
hmm$comp3 <- ifelse(hmm$comp2=="Unknown (S, No Hits)", hmm$comp2, hmm$comp3)

hmm$comp3 <- ifelse(hmm$sporulation_hit!="no_hit", "Sporulation", hmm$comp3)
hmm <- subset(hmm, hmm$comp3=="Sporulation")

hmm <- subset(hmm, hmm$ref=="ref_seq")


hmm.brief <- select(hmm, phage, gene_id, mmseqs_qseqid, predicted_label, detailed, comp, comp2, comp3, spor, Host, life_bool)

hmm.brief <- merge(hmm.brief, dan.enrich, by="gene_id", all.x=TRUE, all.y=FALSE)
hmm.brief$gene_name[is.na(hmm.brief$gene_name)] <- "MISSING"
hmm.brief$gene_name <- ifelse(hmm.brief$gene_name=="MISSING", hmm.brief$gene_id, hmm.brief$gene_name)

### everything is currently sporulating only
#hmm.brief <- unite(hmm.brief, lifestyle_spor, spor, predicted_label, sep = "_", remove=FALSE)
```



```{r}
# add host sporulation to amg
#d.amg <- subset(d.amg, d.amg$phage=="OX639564") ##weird lifestyle data, remove for now
#d.amg <- all %>% rename(spor_host = f_spor, comp3 = comp3, gene_description = query, Lifestyle = predicted_label)



d.amg <- hmm.brief %>% rename(spor_host = Host, Lifestyle = predicted_label)
d.amg %>% group_by(Lifestyle) %>% summarise(n=n())

inph.damg <- d.amg
```


First we summarize the number of phages infecting hosts that are lytic/temp 
```{r}
d.vir <- d.amg %>% group_by(life_bool, Lifestyle) %>% summarise(n=n())

```

```{r}
p.signif <- 1e-3

k_treshold <- 8
```



First we summarize the number of phages infecting hosts that sporulate or not
```{r}

total_temp <- 
  d.vir %>% 
  filter(life_bool) %>% 
  pull(n) 



total_vir <- 
  d.vir %>% 
  filter(!life_bool) %>% 
  pull(n) 


```


For each gene we ask how many times it is found in a temp or lytic phage.
```{r}

# Summarise total number of sporulators and non sporulators
d.amg.sum <- d.amg %>%
  # filter(str_detect(header, regex("sporulation", ignore_case = T))) %>%
  filter(!is.na(gene_name)) %>% 
  filter(!is.na(life_bool)) %>% 
  group_by(gene_name, life_bool) %>% #category removed, lifestyle added
  summarise(n = n()) %>%
  arrange(desc(n))

d.amg.sum %>%
  ggplot(aes(gene_name, n, color = life_bool))+
  geom_point(shape=21, size=2)+
  #facet_wrap(~spor_host, scales = "free_y")+ ## category removed, lifestyle added
  coord_flip()+
  theme_bw()



d.amg.sum$rel <- ifelse(d.amg.sum$life_bool==TRUE, d.amg.sum$n/total_temp*100, d.amg.sum$n/total_vir*100)


rel <- d.amg.sum %>%
  ggplot(aes(gene_name, rel, color = life_bool))+
  geom_point(shape=21, size=2)+
  ylab( "normalized (%) occurence from AMG genes") +
  coord_flip()+ theme_bw() + labs(color="Lifestyle") + scale_color_hue(labels = c("Virulent", "Temperate")) 
  


rel

# ggsave(here("refseq/out-dir/rel_lifestyle_spor_enrich.png"), height =3 ,width = 6)


```



```{r}
temp_enrich <- 
  d.amg %>% 
  
  # Summarize KOs by host sporulation
 mutate(life_bool = case_when(
    life_bool ~ "temperate",
    !life_bool ~ "virulent"
  )) %>% 
  filter(!is.na(gene_name)) %>% 
  group_by(gene_name, life_bool) %>% 
  summarise(n=n(), .groups = "drop") %>% 
  pivot_wider(values_from = n, names_from = life_bool, values_fill = 0) %>% 
   
    # q	: the number of white balls drawn without replacement 
    # from an urn which contains both black and white balls.
    # >>> AMG (specific gene) in viruses infecting sporulators
    mutate(q = temperate) %>% 
    
    # m	:the number of white balls in the urn.
    # >>> number of viruses infecting sporulators in the pool
    mutate(m = total_temp) %>% 
    
    # n	:the number of black balls in the urn.
    # >>> number of viruses infecting non-sporulators in the pool
    mutate(n = total_vir) %>% 
    
    # k: the number of balls drawn from the urn, hence must be in 0,1,., m+n.
    # >>> Total AMG (specific gene) detected in viruses
    mutate(k = virulent + temperate) %>% 
    
    mutate(p.val = signif (phyper(q, m, n, k, lower.tail =F),5)) #%>% 
  # Adjust Pvalue for multiple testing
 #   mutate(p.adj = signif (p.adjust(p.val, method = "BH"),5)) %>% 
  # mark significantly enriched genes
  # mutate(significant = (p.adj < p.signif) & (k>k_treshold))
#pval adjust done later
  
temp_enrich$enriched <- "temperate"
p_temp.all <- temp_enrich
```


```{r}
vir_enrich <- 
  d.amg %>% 
  
  # Summarize KOs by host sporulation
 mutate(life_bool = case_when(
    life_bool ~ "temperate",
    !life_bool ~ "virulent"
  )) %>% 
  filter(!is.na(gene_name)) %>% 
  group_by(gene_name, life_bool) %>% 
  summarise(n=n(), .groups = "drop") %>% 
  pivot_wider(values_from = n, names_from = life_bool, values_fill = 0) %>% 
   
    # q	: the number of white balls drawn without replacement 
    # from an urn which contains both black and white balls.
    # >>> AMG (specific gene) in viruses infecting that lifestyle
    mutate(q = virulent) %>% 
    
    # m	:the number of white balls in the urn.
    # >>> number of viruses of that lifestyle
    mutate(m = total_vir) %>% 
    
    # n	:the number of black balls in the urn.
    # >>> number of viruses infecting the other lifestyle
    mutate(n = total_temp) %>% 
    
    # k: the number of balls drawn from the urn, hence must be in 0,1,., m+n.
    # >>> Total AMG (specific gene) across all lifestyles
    mutate(k = virulent + temperate) %>% 
    
    mutate(p.val = signif (phyper(q, m, n, k, lower.tail =F),5)) #%>% 
  # Adjust Pvalue for multiple testing
    #mutate(p.adj = signif (p.adjust(p.val, method = "BH"),5)) %>% 
  # mark significantly enriched genes
   #mutate(significant = (p.adj < p.signif) & (k>k_treshold))

#pval adjust done later

vir_enrich$enriched <- "virulent"
p_vir.all <- vir_enrich


```

```{r}

## pval adjustement for all?

all.pv <- rbind(p_temp.all, p_vir.all)
# all.pv$og.adj <- all.pv$p.adj


## i adjust p.val original, NOT p.adj (from original?)
  # Adjust Pvalue for multiple testing
all.pv.adj <-  all.pv %>%  mutate(p.adj = signif (p.adjust(p.val, method = "BH"),5)) %>% 
  # mark significantly enriched genes
   mutate(significant = (p.adj < p.signif) & (k>k_treshold))

#all.pv.adj <- merge(all.pv.adj, kegg.enrich, by.x="comp3", by.y="comp3", all.x=TRUE, all.y=FALSE)

##parA is **mostly** in temperate clostridium difficile phage -- think this might be a sampling issue.
```



Plot enrichment results

```{r}
# labels for significant genes
labels <- 
  d.amg %>% 
  select(gene_name) %>% 
  distinct() %>% 
  #filter(comp3 %in% all.pv.adj$comp3[d.vir$significant]) %>% 
  mutate(lab = str_remove(gene_name, "[;|\\[].*")) %>% 
  group_by(gene_name) %>% 
  summarise(p_lab = str_c(lab, collapse = "\n")) #%>% 
  #adjustment of one label
  #mutate(p_lab = if_else(comp3 == "K00558", paste0(p_lab, "\n"), p_lab))

d_enrich <- left_join(all.pv.adj, labels,  by = "gene_name") 


ko.enrich.ids <- read.csv(here(data_dir, "inphared_KO_enrich_hits_eb.csv")) ##d_amg

d_enrich <- merge(d_enrich, ko.enrich.ids, by.x="gene_name", by.y="gene_id", all.x=TRUE, all.y=FALSE)

d_enrich$real <- ifelse(d_enrich$significant==TRUE, d_enrich$gene_name2, NA)


  #mark sporulation genes
d_enrich <- d_enrich %>% 
  mutate(spor_gene = if_else(gene_name %in% unique(d_enrich$gene_name),
         "Spoulation genes", "Other genes"))

```


### commented out code to get unique KO hits and then eb manually rename genes
ko.enrich <- subset(d_enrich, d_enrich$significant==TRUE)
ko.enrich <- ko.enrich[,c(1,14)]
ko.enrich$gene_id <- ko.enrich$gene_name

ko.enrich.kegg <- merge(ko.enrich, kegg.id, by.x="gene_id", by.y="gene_id.ko", all.x=TRUE, all.y=FALSE)

#write.csv(ko.enrich.kegg, here(data_dir, "inphared_KO_enrich_hits.csv"))

ko.enrich.ids <- read.csv(here(data_dir, "inphared_KO_enrich_hits_eb.csv")) ##d_amg

ko.enrich <- merge(ko.enrich, ko.enrich.ids, by="gene_id", all.x=TRUE, all.y=FALSE)





```{r}
# plot
p <-
  d_enrich %>% 
  filter(k>0) %>%
  ggplot(aes(k,-log10(p.adj+1e-120)))+
  geom_hline(yintercept = -log10(p.signif), linetype=2, color="grey", size = 1)+
  geom_vline(xintercept = k_treshold, linetype=2, color="grey", size = 1)+
  
 geom_text_repel(aes(label = real), size = 3, na.rm = T, point.padding = 0, min.segment.length = 0,seed = 123) +
  
  geom_jitter(aes(fill = significant),width = 0.05, height = 0.05,
              shape=21, size=3, stroke = 1, show.legend = F)+
  scale_fill_manual(values=pal2.sail)+
  #scale_fill_viridis_d(direction = -1, alpha = 0.5)+
  theme_classic()+
  facet_wrap(~ enriched %>% fct_rev(), ncol = 2) +
  panel_border(color = "black")+
  scale_x_continuous(trans = "log2", breaks = (2^(0:11)))+
  xlab("Sample size\nNo. homologs detected (log2)")+
  ylab("Enrichment (-log10 adj. P-value)")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

p

ggsave(here(figs_dir, "inph_lifestyle_SPORE_enriched.png"), height =4 ,width = 6)
 #ggsave(here("mmseqs/inphared/inph_lifestyle_spor_enrich.png"), height =4 ,width = 6)

```



```{r}
sig <- subset(all.pv.adj, all.pv.adj$significant==TRUE)

sig.list <- sig[,c(1,9)]
sig.list$life_bool <- TRUE
sig.list2 <- sig[,c(1,9)]
sig.list2$life_bool <- FALSE
sig.list <- rbind(sig.list, sig.list2)




# Summarise total number of sporulators and non sporulators
d.amg.sum <- d.amg %>%
  # filter(str_detect(header, regex("sporulation", ignore_case = T))) %>%
  filter(!is.na(gene_name)) %>% 
  filter(!is.na(life_bool)) %>% 
  group_by(gene_name, life_bool) %>% #category removed, lifestyle added
  summarise(n = n()) %>%
  arrange(desc(n))

d.amg.sum %>%
  ggplot(aes(gene_name, n, color = life_bool))+
  geom_point(shape=21, size=2)+
  #facet_wrap(~spor_host, scales = "free_y")+ ## category removed, lifestyle added
  coord_flip()+
  theme_bw()



d.amg.sum$rel <- ifelse(d.amg.sum$life_bool==TRUE, d.amg.sum$n/total_temp*100, d.amg.sum$n/total_vir*100)

d.amg.sig <- sig.list %>%
  left_join(d.amg.sum, by = c("gene_name", "life_bool"))



d.amg.sig <- merge(d.amg.sig, ko.enrich.ids, by.x="gene_name", by.y="gene_id", all.x=TRUE, all.y=FALSE)


#d.amg.sig$rel[is.na(d.amg.sig$rel)] <- 0

d.amg.sig <- d.amg.sig %>%
  mutate(enriched_facet = fct_rev(enriched))


d.amg.sig <- d.amg.sig %>%
  group_by(enriched_facet) %>%
  mutate(gene_name = fct_drop(factor(gene_name))) %>%
  ungroup()


rel <- d.amg.sig %>%
  ggplot(aes(gene_name2, rel, fill= life_bool)) +
  #geom_point(shape = 21, size = 2) +
  geom_bar(stat="identity", position=position_dodge()) +
  ylab("Normalized (%) Relative Abundance of Gene") +
  theme_bw() +
  labs(color = "Lifestyle", fill = "Phage Lifestyle") +
  scale_color_hue(labels = c("Virulent", "Temperate")) +
  facet_wrap(~ enriched_facet, scales = "free_x", ncol = 2) + scale_fill_manual(labels = c ("Lytic", "Temperate"), values=pal2.flip) +theme(axis.text.x = element_text(angle = 25, hjust = 1)) + xlab("Enriched Gene") + ggtitle("INPHARED") +theme(legend.position = "bottom") #+ coord_flip()


rel
ggsave(here(figs_dir, "inph_lifestyle_SPORE_enriched_relative.png"), height =4 ,width = 6)
 #ggsave(here("refseq/out-dir/rel_refseq_lifestyle_spor_enrich.png"), height =4 ,width = 4.5)

#geom_boxplot(binaxis = "y", stackdir = "center", position = "dodge") + geom_jitter(width = 0.2) + coord_flip()

```






```{r}
d.amg.sig.bubble <- d.amg.sig
d.amg.sig.bubble$life_bool <- ifelse(d.amg.sig.bubble$life_bool==TRUE, "Temperate", "Virulent")
d.amg.sig.bubble$enriched_facet <- ifelse(d.amg.sig.bubble$enriched_facet=="temperate", "Temperate-Enriched Genes", "Virulent-Enriched Genes")
#d.amg.sig.bubble <- subset(d.amg.sig, d.amg.sig$enriched_facet=="temperate")
#d.amg.sig.bubble <- d.amg.sig.bubble %>%
#  mutate(comp3 = factor(comp3, levels = pathway_levels))
list <- unique(select(d.amg.sig.bubble, gene_name2, enriched_facet))
#write.csv(list, here(figs_dir, "inph_spor_enrich_sig.csv"))
enrich<- read.csv(here(figs_dir, "inph_spor_enrich_sig_EB.csv"))

d.amg.sig.bubble <- d.amg.sig.bubble %>%
  full_join(enrich, by = c("gene_name2"))


pathway_levels <- d.amg.sig.bubble %>%
  distinct(Stage, gene_name2) %>%
  arrange(Stage, gene_name2) %>%
  pull(gene_name2)


d.amg.sig.bubble$gene_name2 <- factor(d.amg.sig.bubble$gene_name2, levels = pathway_levels)



```




```{r}
size_min_observed <- min(d.amg.sig.bubble$rel[d.amg.sig.bubble$rel > 0], na.rm = TRUE) # e.g. 0.1
size_limits <- c(0.05, 25)
size_max    <- 8
size_breaks <- c(0.05, 1, 5, 10,20)
size_labels <- as.character(size_breaks)

p <- ggplot(d.amg.sig.bubble, aes(x = life_bool, y = gene_name2)) +
  geom_point(aes(size = rel), shape = 21, stroke = 0.6, fill="black") +
 scale_size_area(limits = size_limits, breaks = size_breaks, labels = size_labels, max_size = size_max)+
  # control max bubble diameter
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(size = 11, face = "bold"),
    axis.text.y = element_text(size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ) +facet_wrap(~ enriched_facet, scales = "free_y", ncol = 2)  

p

# Add a colored strip at the LEFT that corresponds to pathway categories
strip_df <- unique(select(d.amg.sig.bubble, enriched_facet, gene_name2, Stage))
strip_df$Stage <- as.factor(strip_df$Stage)
strip_df <- strip_df %>% mutate(xpos = as.numeric(Stage))

strip_x <- min(strip_df$xpos) - 0.8   # ← ONLY CHANGE (was max + offset)

strip_df <- strip_df %>%
  select(gene_name2, Stage, enriched_facet) %>%
  distinct() %>%
  mutate(x = strip_x, xend = strip_x + 0.4)

p <- p +
  geom_tile(
    data = strip_df,
    aes(x = x, y = gene_name2, fill = Stage),
    width = 0.4,
    height = 0.9,
    inherit.aes = FALSE
  ) +
  coord_cartesian(clip = "off")

# Add legend and show
temp.gg <- p + guides(size = guide_legend(title = "Gene Category Proportion (%) "), fill=guide_legend("Sporulation Function"))
temp.gg

ggsave(here(figs_dir, "inph_SPOR_bubbleplot_Enrich_Stage.png"), height =6 ,width = 14)

```