---
title: "R Notebook"
output: html_notebook
---

```{r}
library(here)
library(tidyverse)
library(PNWColors)
library(knitr)
library(cowplot)




kegg <- read.csv(here("00_data", "03_func_data", "func_metadata", "viral_sporGenes.csv"))
kegg$source <- "KEGG"

kegg$gene_name <- gsub(".*_(.*)$", "\\1", kegg$gene_name)
kegg$gene_name <- ifelse(kegg$gene_id=="K03657", "pcrA", kegg$gene_name)
kegg$description[is.na(kegg$description)] <- kegg$gene_name

kegg <- select(kegg, gene_id, gene_name, description, source)

colnames(kegg) <- c("ID", "gene_name", "gene_desc", "source")
```
```{r}
COG <- read.csv(here("00_data", "03_func_data", "func_metadata", "COG", "COG_sporList.csv"),row.names=1) 
COG$source <- "COG"

###for some reaason, COG5928 is just not named in COG db
COG$gene_name <- ifelse(COG$COG=="COG5928", "SpoIIT/YwmB", COG$gene_name)
COG$gene_desc <- ifelse(COG$COG=="COG5928", "Sporulation protein SpoIIT/YwmB, DUF1779 family", COG$gene_desc)
COG$Cat <- ifelse(COG$COG=="COG5928", "D", COG$Cat)

COG <- na.omit(COG)
COG$gene_name <- ifelse(COG$gene_name=="", COG$gene_desc, COG$gene_name)
duplicate_rows <- COG[duplicated(COG$gene_name), ]
COG <- select(COG, COG, gene_name, gene_desc, source)

colnames(COG) <- c("ID", "gene_name", "gene_desc", "source")
```


```{r}
#P29600
DRAM <- read.csv(here("00_data", "03_func_data", "func_metadata",  "UniprotSporGerm_Master.csv"),row.names=1)
DRAM$source <- "DRAM"


DRAM <-select(DRAM, Entry, Gene.Names, Protein.names, gene_name, source, Pfam)
DRAM[is.na(DRAM)] <- 0

DRAM<- separate(DRAM, Gene.Names, into=c("FirstGene", "Extra"), sep=" ", remove=FALSE, extra="merge")
DRAM$gene_name <- ifelse(DRAM$gene_name==0 | DRAM$gene_name=="", DRAM$FirstGene, DRAM$gene_name)

DRAM$gene_name <- ifelse(DRAM$gene_name=="", DRAM$Protein.names, DRAM$gene_name)


DRAM <- select(DRAM, Entry, gene_name, Protein.names, source)

colnames(DRAM) <- c("ID", "gene_name", "gene_desc", "source")


```


###originally to clean phrogs

library(here)
library(dplyr)
library(readr)

# path to the folder (relative to project root)
csv_dir <- here("00_data", "03_func_data", "func_metadata",  "PHROGS")

# list all csv files
csv_files <- list.files(
  csv_dir,
  pattern = "\\.csv$",
  full.names = TRUE
)

# read and concatenate
phrogs <- csv_files %>%
  lapply(read_csv, show_col_types = FALSE) %>%
  bind_rows()


phrogs <- unique(phrogs)
phrogs <- separate(phrogs, KEGG, into=c("KO", "KO_name", "KO_ignore"), sep="&", remove=FALSE, extra="merge")
phrogs <- separate(phrogs, pfam, into=c("FirstPfam", "2ndPfam"), sep="\\.", remove=FALSE, extra="merge")
phrogs <- phrogs[,c()]
phrogs[is.na(phrogs)] <- "NA"

phrogs$gene_name <- phrogs$KO_name
phrogs$gene_name <- ifelse(phrogs$gene_name=="NA", phrogs$Phrog_annot, phrogs$gene_name)


#write.csv(phrogs, here("00_data", "03_func_data", "func_metadata", "PHROGS", "manualCleaning_phrogs_IDs.csv"))

```{r}

phrogs.man <- read.csv(here("00_data", "03_func_data", "func_metadata", "PHROGS", "manualCleaning_phrogs_IDs_EB.csv"))

phrogs.man <- subset(phrogs.man, phrogs.man$KEEP=="Y")
phrogs.man <- select(phrogs.man, X.phrog, gene_name_final, KEGG)
phrogs.man$source <- "PHROGS"

colnames(phrogs.man) <- c("ID", "gene_name", "gene_desc", "source")


```



```{r}
all.spor.prots <- rbind(kegg, COG, DRAM, phrogs.man)


colnames(all.spor.prots) <- c("ID", "gene_name", "gene_desc", "best_db")
all.spor.prots$sporprot <- "sporulation_prot"
### to make gene_names consistent, lowercase the first lower (so clwJ and ClwJ both match)
## this may be grammatically incorrect for some, but prevents duplicate categories 

all.spor.prots$gene_name <- str_replace(
  all.spor.prots$gene_name,
  "^.",
  tolower
)

write.csv(all.spor.prots, here("00_data", "03_func_data", "func_metadata", "all_sporProt_IDs.csv"))

```

